---
title: "STAT 5301 SP2020 Homework 1 by Curtis Isaacson.32"
output:
  word_document: default
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---


 HOMEWORK QUESTIONS 


After each of the questions below, write some R code that does the required task
and run the R code.  FIRST, insert the necessary chunk of R Markdown code that is needed to run the Rcode that you will be writing.  Select the Insert button to the left of the Run button above and select R.  You will get something like the following:
 

'```{r}'

'```'

Insert your R code between the lines containing ```.  This tells R Markdown to expect R code when "Knitting" (or compiling) the final output document.  Click the Insert button at the top and Select the R option from the list.  This will insert the following:

```{r}

```


Make sure to run each chunk of R code so that output is produced. To run the code, put the cursor within the desired R chunk of code and select the Run button and then Select Run Current Chunk.   Or click on the green arrow on the right side of the code that is pointing to the right (not downward).  This will run the particular chunk of code in this section.  Select the downward arrow to run all chunks of R code.   This needs to be done at least once, otherwise the R code appears in the resulting output document after you "knit" everything together rather than the output the code generates. 

To see how the final output will appear, select the arrow next to the Knit button (or Preview button) and select Knit to Word.  This will create a Word Dcoument called Homework1-SP20.docx in the folder where you stored Homework1-SP20.Rmd.  To create a PDF version of your final output, knit your document as a Word or HTML document and select Print to PDF or Save as PDF.  Upload the final PDF version into Carmen.
 
Optional:  If you successfully installed TinyTex (or MikTex), select "knit to PDF" which will create a new line to your header under "output" at the top of your document:  pdf_document: default.  Select "knit to PDF"  again and a PDF file called Homework1-3.pdf should be created. Upload the final PDF document to Carmen.  Note:  you have to close the .pdf viewer, which displays the resulting .pdf file, before you are able to run "Knit to PDF" successfully again.  It will result in an error.

This homework is being graded for attempts, rather than accuracy.  Refer to the lecture notes for examples of R code. Additional help is provided by the links in the Installation of R and RStudio document.  When the assignement is completed, please upload two files, your .RMD file and the output as a .pdf file. Filenames for each homework assignment will have the following format:  lastname_firstname_hw#.Rmd which contains the text and R code, and lastname_firstname_hw#.pdf, which is a copy of the output generated by knitting lastname_firstname_hw#.Rmd.  For example, for homework 1, I would submit kelbick_nicole_hw1.Rmd and kelbick_nicole_hw1.pdf.

Good luck!


HOMEWORK 1

 1. Define a new variable, 'monkey', that contains the following sequence of
    integers ranging from -100 to 50 and increasing by 5 (e.g. -100 -95 ... 45 50).
    Display this variable.

```{r}
monkey = seq(-100, 50, by = 5)

```


 2. a. Check whether each entry of 'monkey' is negative and display results.
       Results will be a vector of TRUE and FALSE
       
```{r}

```


    b. Determine how many values in this sequence of numbers are negative.

 3. In the "datasets" package there is a data set called 'trees' that
    provides information on the girth (circumference), height, and volume of 31 
    black cherry trees.
    a. Load this data set into R.

    b. Display the first few lines of the data set.

    c. Plot tree height (on the x axis) vs. volume (on the y axis)
       MAKE SURE TO: label the x and y axes with the variable name AND the units
       It will be helpful to look at help(trees) and help(plot).

    d. Repeat part c but replace height with girth.

    e. Which variable, height or girth, does a better job at predicting volume.  
       Explain briefly.

 4. Continuing to work with the 'trees' data:
    a. Define a vector 'under11' which is a vector of length 31 and is TRUE
       if the girth of the tree is less than or equal to 11 inches.

    b. Run the code: 'cbind(trees, under11)' to verify that you have succeeded.  
       Just show the resulting output.  
       (NOTE:  cbind() will add under11 as a column to the trees data frame.  You
       will have succeed if this command runs without an error.  rbind() works in
       a similar fashion but adds rows to an existing data frame rather than 
       columns.)

    c. Evaluate 'length(under11)' and 'sum(under11)'.  What do these numbers tell 
       you?

 5. Continuing to work with the 'trees' data:
    a. Define a new data set, 'skinny.trees', which contains only information on 
       the trees whose girth does not exceed 11 inches.  (It will be a subset of 
       the original trees dataset.)

    b. Plot tree girth vs. volume for just these trees.

    c. Make the points with a girth of 11 inches be filled in and blue in color.
       NOTE: There are a variety of ways to do this but you could try to use the 
       'which' command to help you -- try 'help(which)' to get started.




